---
title: "Data frames: Your new favorite data structure"
author: ~
date: '2020-09-10'
slug: "data-frames"
categories: [notes]
tags: [week04]
draft: false
# toc: true
---

Discusses the data structure called data frames: what they are and why they are so important.

<!--more-->

```{r setup, echo=TRUE, message=FALSE}
#library(tidyverse)
#knitr::opts_chunk$set(eval=TRUE)
```

# Review of data structures

In lecture on Wednesday, Sept 9, we discussed a key data structure in R: **vectors**^[Properly speaking, these are called atomic vectors. However, colloquially we refer to these as vectors.]. There are three important rules for vectors:

* They are **1-dimensional**: think of them like a column of values.
* They are **ordered**: there is a first element, a second element, etc.
* They contain **a single data type**: a vector can be integer, double, logical, or character, but they can't be a mix of those. If a vector looks like it has a mix of data types, that's because it has been *coerced* to a more general type. For instance, `c(7, "aardvark")` will be coerced to a character vector: the number `7` will be converted to the character `"7"`.

There are a two other data structures which are frequently used in base R, but not in the tidyverse:

* **Matrix**: a 2-dimensional vector. This follows the same rules as vectors, but it has rows and columns.
* **List**: I think of this like a bag, into which you can throw any data structure you want. (Lists are ordered, like vectors, so maybe it is an unusually well-ordered bag?) The first element of a list might be a character vector, the second might be a integer matrix, and the third might be another list (which might, in turn, contain more lists)! These are great when you have to keep track of non-parallel information about something. For instance, a linear model object contains the coefficients of the linear model,  the residuals, and all sorts of other information that comes in different data types and different lengths. The `lm()` function therefore returns a special kind of list to store all that information.

# Data frames and tibbles

A data frame is a special kind of list. Data frames must obey two rules:

* The list can only contain **vectors**
* Each vector must be the same **length**

Therefore, data frames act like a rectangular table. For instance, here are  the first six rows of `mtcars`, a data frame  that is built into R's **datasets** package, which is autoloaded when you start R:

```{r, echo = TRUE, eval = TRUE}
head(mtcars) # head() prints the first 6 rows of adata frame
```

## Creating data frames

We can create data frames manually from vectors using the `data.frame()` function. Recall that we can use `c()` (concatenate) to build vectors. We name the columns of the data frame using the arguments of the function: in the example below `x.col = x` says "take the vector x, and name it `x.col` in the new data frame.

```{r}
x <- c(1, 3, 5)
y <- c("a", "c", "d")
d <- data.frame(x.col = x, y.col = y)
d
```

`d` looks like a data frame when it is printed out, but we really should check to be sure:

```{r}
str(d)
```

Note that `obs.` (short for observations) corresponds to 'rows' and `variables` corresponds to 'columns'. 

Of course, it is pretty rare that we want to create data frames by hand^[Usually I do this when I am making a reproducible example, or when I am generating some example data to test some code]. More often, we read in a file, for instance a text file. The functions in the **readr** package like `read_csv()` (which reads a comma-separated value file).

# Data frames vs tibbles

The tidyverse is built around data frames. Basically, tidyverse wants you to use data frames whenever possible. Because data frames are so important, tidyverse defines a special kind of data frame: the **tibble**. (The name comes the function `tbl_df`, or table-data-frame). 

For our purposes, tibbles are equivalent to data frames. Tibbles contain a little more information when printed out, and protect you a little more when you do something silly to them. But for our purposes, whenever you hear "tibble", you can think "data frame". 


